---
title: "Analysis_1"
author: "Nataraj.Patankar.Tiwari.Wavre"
date: "2025-04-16"
output: pdf_document
editor_options: 
  chunk_output_type: console
---

## Number of EVs vs Number of Charging Stations

In this part of our analysis, we explore  

```{r analysis 1 part 1}
#Loading Packages
```

## Number of EVs vs Average income of States

In this part of our analysis, we explore whether the average income of a state is correlated to the number of electric vehicles sold in that state. To do so, we will start by cleaning our data, conducting a correlation test, and then creating a scatter plot with a linear regression line. 

```{r analysis 1 part 2, warning=FALSE, error=FALSE, message=FALSE}
#Loading Packages 
library(dplyr)
library(tidyverse)
library(here)
library(ggplot2)
library(corrplot)
library(readxl)
library(stringr)

here()

#Cleaning Data
EV_Data_PT <- read_excel("Data/Raw/2016-2023 LDV Registration Data.xlsx")
Income_Data_PT <- read.csv("Data/Raw/Average Income Data by State.csv")
Income_Clean <- Income_Data_PT %>%
  mutate(State = str_trim(str_remove(State, "\\*")))

EV_vs_Income <- left_join(EV_Data_PT, Income_Clean, by = c("State", "Year"))
colnames(EV_vs_Income) [c(3,4,15)] <- c("EV", "PHEV","Avg_Income")

EV_vs_Inc_clean <- EV_vs_Income %>%
  select(Year, State, EV, PHEV, Avg_Income) %>%
  mutate(Total_EV = EV + PHEV)

#Correlation test
EV.test <- cor.test(EV_vs_Inc_clean$Total_EV, EV_vs_Inc_clean$Avg_Income)
EV.test

#Plotting the data
ggplot(EV_vs_Inc_clean, aes(x = Avg_Income, y = Total_EV)) +
  geom_point(alpha = 0.6, color = "steelblue") +
  geom_smooth(method = "lm", se = TRUE, color = "darkred", linewidth = 1.2) +
  labs(
    title = "Relationship Between Average Income and EV Registrations",
    subtitle = "Across States (Total EVs = EV + PHEV)",
    x = "Average Income (USD)",
    y = "Total EV Registrations"
  ) +
  ylim(0,250000) +
  theme_minimal()

```

## Number of EVs vs Population of States

In this part of our analysis, we explore whether the population of a state is correlated to the number of electric vehicles sold in that state. To do so, we will start by cleaning our data, conducting a correlation test, and then creating a scatter plot with a linear regression line. 

```{r analysis 1 part 3}
#Cleaning Data
EV_Data_PT <- read_excel("Data/Raw/2016-2023 LDV Registration Data.xlsx")
Pop_Data_PT <- read_excel("Data/Raw/Population 2020-2024.xlsx")
colnames(Pop_Data_PT) [c(2,3,4,5)] <- c("Pop_2020","Pop_2021","Pop_2022","Pop_2023")

Pop_clean <- Pop_Data_PT %>%
  pivot_longer(cols = starts_with("Pop_"), names_to = "Year", names_prefix = "Pop_", values_to = "Population") %>%
  select(Year, State, Population) %>%
  mutate(Year = as.numeric(Year))
  
EV_vs_Pop <- left_join(EV_Data_PT, Pop_clean, by = c("Year", "State"))
colnames(EV_vs_Pop) [c(3,4)] <- c("EV", "PHEV")

EV_vs_Pop_clean <- EV_vs_Pop %>%
  select(Year, State, EV, PHEV, Population) %>%
   mutate(Total_EV = EV + PHEV)

#Correlation test
Pop.test <- cor.test(EV_vs_Pop_clean$Total_EV, EV_vs_Pop_clean$Population)
Pop.test

#Plotting the data
ggplot(EV_vs_Pop_clean, aes(x = Population, y = Total_EV)) +
  geom_point(alpha = 0.8, color = "coral") +
  geom_smooth(method = "lm", se = TRUE, color = "navy", linewidth = 1.2) +
  labs(
    title = "Relationship Between Population and EV Registrations",
    subtitle = "Across States (Total EVs = EV + PHEV)",
    x = "Population",
    y = "Total EV Registrations"
  ) +
   ylim(0,250000) +
  xlim(0,30000000) +
  theme_minimal()

```

Note that the `echo = FALSE` parameter was added to the code chunk to prevent printing of the R code that generated the plot.
