---
title: "Factors Affecting EV Adoption in the U.S."
author: "H.Nataraj, A.Patankar, P.Tiwari, P.Wavre"
date: "24 April 2025"
output: 
  html_document:
    toc: true
editor_options: 
  chunk_output_type: console
---

## Introduction
Electric vehicles (EVs) are rapidly transforming the U.S. transportation landscape, driven by growing concerns over climate change, rising fuel prices, and supportive policy incentives. As the market expands, understanding the factors that are influencing the adoption of EVs is critical for designing equitable and effective transportation policies.

## Background
The global transition toward EVs is accelerating, driven by government policies, technological advancements, and increasing environmental awareness. Alongside these developments, fueling infrastructure in the U.S. continues to play a critical role, both in supporting traditional internal combustion engine (ICE) vehicles and facilitating the growth of the EV market. Additionally, as EVs remain relatively new and expensive technology, areas with higher purchasing power, larger populations, and greater access to charging infrastructure may experience faster and more substantial adoption rates.

## Motivation
Analyzing the relationship between EV sales trends and fuel station infrastructure is essential for effective infrastructure planning, informed policymaking, and targeted investment strategies. As EV adoption accelerates, the demand for traditional gas stations is expected to decline, accompanied by a corresponding rise in the need for robust EV charging infrastructure. Understanding these evolving trends can guide strategic decisions on how the U.S. can best adapt its fueling networks to support the clean energy transition. Furthermore, examining demographic factors such as population density and income levels can help determine optimal locations for deploying EV charging stations.

## Information on Datasets
We used five primary datasets in this analysis:
```{r datasets table, message=FALSE, warning=FALSE, echo=FALSE}
#Dataset info
dataset_info <- data.frame(
  Dataset = c("Vehicle Registration Counts", 
              "Average Income by State", 
              "Population",
              "Alternative Fueling Station Counts",
              "Cartographic Boundary Shapefiles"),
  Source = c("[Alternative Fuels Data Center](https://afdc.energy.gov/vehicle-registration)",
             "[Bureau of Economic Analysis](https://apps.bea.gov/itable/?ReqID=70&step=1&_gl=1*183o7fy*_ga*MTA2NTA5NzQzNS4xNzQ0NzQ1MTM1*_ga_J4698JNNFT*MTc0NTAwNzIyOS4zLjAuMTc0NTAwNzIyOS42MC4wLjA.#eyJhcHBpZCI6NzAsInN0ZXBzIjpbMSwyOV0sImRhdGEiOltbIlRhYmxlSWQiLCIyMSJdXX0=)",
             "[U.S. Census Bureau](https://www.census.gov/data/tables/time-series/demo/popest/2020s-state-total.html)",
             "[Alternative Fuels Data Center](https://afdc.energy.gov/stations/states)",
             "[U.S. Census Bureau](https://www.census.gov/geographies/mapping-files/time-series/geo/carto-boundary-file.html)"),
  Years = c("2016–2023", "2016–2023", "2020–2024","2016-2023","NA")
)

#Converting dataset info into a table
library(knitr)
kable(dataset_info, format = "markdown")

#1. The [Energy Efficiency and Renewable Energy's Alternative Fuels Dataset](https://afdc.energy.gov/stations/states) from U.S. Department of Energy provided alternative fueling station counts by state and fuel type, along with the vehicle registration counts.
#2. Income data was sourced from the [U.S. Bureau of Economic Analysis](), which reports annual per capita personal income.
#3. Population estimates were taken from the [U.S. Census Bureau](), covering state-level populations for each year in the study period.
#4. The U.S. Census Bureau also provided the [Cartographic Boundary Shapefiles]() for all the states-5m.
```

## Setting up the Session
```{r setup, message=FALSE, warning=FALSE, results='hide'}
#Installing and loading packages
#install.packages("ggforce")
#install.packages("readxl")
library(tidyverse)
library(lubridate)
library(here)
library(ggplot2)
library(stringr)
library(cowplot)
library(ggforce)
library(scales)
library(dplyr)
library(readxl)
library(sf)
library(leaflet)
library(mapview); mapviewOptions(fgb = FALSE)

#Check working directory
here()
```

## Importing and Cleaning Datasets

```{r datasets, message=FALSE, warning=FALSE, results='hide'}
#1. EV Registration Data
LDV_Data <- read_excel("Data/Raw/2016-2023 LDV Registration Data.xlsx")

#Renaming columns 
colnames(LDV_Data) [c(3,4)] <- c("EV", "PHEV")

#Selecting and creating columns
LDV_Clean <- LDV_Data %>%
  mutate(Total_EV = EV + PHEV) %>%
  select(Year, State, Total_EV, EV, PHEV)

#write.csv(LDV_Clean, "Data/Processed/LDV_Clean.csv", row.names = FALSE)

#2. Income Data
Income_Data <- read.csv("Data/Raw/Average Income Data by State.csv")

#Cleaning data
Income_Clean <- Income_Data %>%
  mutate(State = str_trim(str_remove(State, "\\*"))) %>%
  rename("Avg_Income" = "Per.capita.personal.income..dollars.")

#3. Population Data
Pop_Data <- read_excel("Data/Raw/Population 2020-2024.xlsx")

#Renaming the columns
colnames(Pop_Data) [c(2,3,4,5)] <- c("Pop_2020","Pop_2021","Pop_2022","Pop_2023")

#Wrangling the data
Pop_Clean <- Pop_Data %>%
  pivot_longer(cols = starts_with("Pop_"), names_to = "Year", names_prefix = "Pop_", values_to = "Population") %>%
  select(Year, State, Population) %>%
  mutate(Year = as.numeric(Year))

#write.csv(Pop_clean, "Data/Processed/Population_Clean.csv", row.names = FALSE)

#4. Fuel Stations 
Fuel_Stations <- read.csv(file = here("Data/Raw/Fuel_Stations_Historical_Data.csv"),
stringsAsFactors = TRUE)

#Selecting columns
Fuel_Stations <- Fuel_Stations[, c("Station.Name", "State", "Latitude", "Longitude", "Open.Date")]

#Formatting dates
Fuel_Stations$Open.Date <- dmy(Fuel_Stations$Open.Date)
Fuel_Stations$Open.Date <- year(Fuel_Stations$Open.Date)
Fuel_Stations <- subset(Fuel_Stations, Open.Date >= 2016 & Open.Date <= 2023)

#Converting acronyms to state names 
Fuel_Stations$State <- toupper(trimws(Fuel_Stations$State)) 

state_abbr <- state.abb 
state_names <- state.name 

state_abbr <- c(state_abbr, "DC")
state_names <- c(state_names, "District of Columbia") #Added DC manually

Fuel_Stations$State <- state_names[match(Fuel_Stations$State, state_abbr)] 

#Regrouping the stations 
Fuel_Stations_Summary <- Fuel_Stations %>%
  group_by(Open.Date, State) %>%
  summarise(Number_of_Stations = n(), .groups = 'drop') %>%
  rename(Year = Open.Date)

#write.csv(Fuel_Stations_Summary, "Data/Processed/Fuel_Stations_Summary.csv", row.names = FALSE)

#Data for Maps
States_sf <- st_read( here('Data/Raw/cb_2018_us_state_5m/cb_2018_us_state_5m.shp'))
Stations_Data <- read.csv(file = here("Data/Raw/Fuel_Stations_Historical_Data.csv"),
stringsAsFactors = TRUE)
```

## Analysis 0: Visualizing Data through Graphs
### Summary Plots for EV Registration Data
Explain what we're doing here for each graph

```{r summary plots, message=FALSE, warning=FALSE, fig.width=14, fig.height=8}
#Plots of EV registrations
ggplot(LDV_Clean, aes(x = Year, y = `Total_EV`, color = State)) +
  geom_line(size = 1) +
  labs(
    title = "EV Registrations by State (2016–2023)",
    x = "Year",
    y = "Number of EVs Registered",
    color = "State"
  ) +
  theme_minimal(base_size = 14)

#Filtering the dataset for California
LDV_Clean %>%
  filter(State != "California") %>%
  ggplot(aes(x = Year, y = `Total_EV`, color = State)) +
  geom_line(size = 1) +
  labs(
    title = "EV Registrations by State (Excl. California, 2016–2023)",
    x = "Year",
    y = "Number of EVs Registered",
    color = "State"
  ) +
  theme_minimal(base_size = 14)

```

``` {r california plot, message=FALSE, warning=FALSE}
#Plotting California separately 
ggplot(filter(LDV_Clean, State == "California"), 
       aes(x = Year, y = `Total_EV`)) +
  geom_line(color = "navy", size = 1) +
  labs(
    title = "EV Registrations in California (2016–2023)",
    x = "Year", y = "EV Registrations"
  ) +
  theme_minimal(base_size = 12)

```

> Analysis: We can see that 

### Maps for Charging Stations Data
Explain what we're doing here (Creating a map of Charging Stations all of US - Includes Mainland USA, Alaska and Hawaii)

```{r maps, message=FALSE, warning=FALSE}
#
bbox_us <- st_bbox(c(xmin = -179, xmax = -66, ymin = 18.5, ymax = 72), crs = st_crs(States_sf))
States_sf_cropped <- st_crop(States_sf, bbox_us)

#
mapviewOptions(basemaps = "OpenStreetMap")
#mapview(States_sf_cropped)

#Selecting required columns and removing rows which do not have valid latitude and longitude data
Stations_Map <- Stations_Data %>%
  filter(Fuel.Type.Code == "ELEC") %>%
  select(Fuel.Type.Code, Latitude, Longitude) %>%
  filter(
    !is.na(Longitude), !is.na(Latitude),
    Latitude >= 18.5, Latitude <= 72,
    Longitude <= -66, Longitude >= -179)

#write.csv(Stations_Map, "Data/Processed/EVStation_Locations.csv", row.names = FALSE)

#
Stations.sf <- Stations_Map %>%
  st_as_sf(
    coords = c('Longitude','Latitude'),
    crs=4269)

#Mapping the data
Plot_EVStationLocations <- ggplot()+
  geom_sf(data = States_sf, fill="white", size = 0.5) +
  geom_sf(data = Stations.sf, color = "red", size = 0.2)+
  ggtitle("EV Station Locations in the U.S.")+
  coord_sf(
    xlim = c(-179, -66), 
    ylim = c(18.5, 72),    
    expand = FALSE)
Plot_EVStationLocations
```

### Map of Mainland U.S.
Write one line about why

```{r clean map, message=FALSE, warning=FALSE}
#
bbox_us2 <- st_bbox(c(xmin = -125, xmax = -66, ymin = 24, ymax = 50), crs = st_crs(States_sf))
States_sf_cropped2 <- st_crop(States_sf, bbox_us2)
#mapview(States_sf_cropped2)

#Selecting required columns and removing rows which do not have valid latitude and longitude data
Mainland_Map <- Stations_Data %>%
  filter(Fuel.Type.Code == "ELEC") %>%
  select(Fuel.Type.Code, Latitude, Longitude) %>%
  filter(
    !is.na(Longitude), !is.na(Latitude),
    Latitude >= 24, Latitude <= 50,
    Longitude <= -66, Longitude >= -125)

#
Stations.sf2 <- Mainland_Map %>%
  st_as_sf(
    coords = c('Longitude','Latitude'),
    crs=4269)

#Mapping the data
Plot_EVStationLocations2 <- ggplot()+
  geom_sf(data = States_sf, fill="white", size = 0.5) +
  geom_sf(data = Stations.sf2, color = "red", size = 0.2)+
  ggtitle("EV Station Locations in the Mainland U.S.")+
  coord_sf(
    xlim = c(-125, -66), 
    ylim = c(24, 50),    
    expand = FALSE)
Plot_EVStationLocations2
```

> Analysis: we see that 

\newpage

## Analysis 1: Effect of Various Factors on EV Sales
### Number of EVs vs Number of Charging Stations

In this part of our analysis, we explore  

```{r analysis 1 part 1}
#Loading Packages
```

### Number of EVs vs Average income of States

In this part of our analysis, we explore whether the average income of a state is correlated to the number of electric vehicles sold in that state. To do so, we will start by cleaning our data, conducting a correlation test, and then creating a scatter plot with a linear regression line. 

```{r analysis 1 part 2, warning=FALSE, error=FALSE, message=FALSE}

#Wrangling Data
EV_vs_Income <- left_join(LDV_Clean, Income_Clean, by = c("State", "Year"))

#Correlation test
EV.test <- cor.test(EV_vs_Income$Total_EV, EV_vs_Income$Avg_Income)
EV.test

#Plotting the data
ggplot(EV_vs_Income, aes(x = Avg_Income, y = Total_EV)) +
  geom_point(alpha = 0.6, color = "purple") +
  geom_smooth(method = "lm", se = TRUE, color = "darkgreen", linewidth = 1.2) +
  labs(
    title = "Relationship Between Average Income and EV Registrations",
    subtitle = "Across States (Total EVs = EV + PHEV)",
    x = "Average Income (USD)",
    y = "Total EV Registrations"
  ) +
  ylim(0,250000) +
  theme_minimal()

```

> Analysis: We observe a moderate positive correlation between average income and EV registrations, with a Pearson correlation coefficient of 0.28 (p ≈ 4.83e-09). Although not as strong as the population relationship, this still indicates that higher-income states are more likely to have greater EV uptake. This aligns with expectations, as electric vehicles are often priced higher than conventional cars, and wealthier populations may have more disposable income and access to incentives for EV purchases.

### Number of EVs vs Population of States

In this part of our analysis, we explore whether the population of a state is correlated to the number of electric vehicles sold in that state. To do so, we will start by cleaning our data, conducting a correlation test, and then creating a scatter plot with a linear regression line. 

```{r analysis 1 part 3, warning=FALSE, error=FALSE, message=FALSE}
#Wrangling Data
EV_vs_Pop <- left_join(LDV_Clean, Pop_Clean, by = c("Year", "State"))

#Correlation test
Pop.test <- cor.test(EV_vs_Pop$Total_EV, EV_vs_Pop$Population)
Pop.test

#Plotting the data
ggplot(EV_vs_Pop, aes(x = Population, y = Total_EV)) +
  geom_point(alpha = 0.8, color = "coral") +
  geom_smooth(method = "lm", se = TRUE, color = "navy", linewidth = 1.2) +
  labs(
    title = "Relationship Between Population and EV Registrations",
    subtitle = "Across States (Total EVs = EV + PHEV)",
    x = "Population",
    y = "Total EV Registrations"
  ) +
   ylim(0,250000) +
  xlim(0,30000000) +
  theme_minimal()

```

> Analysis: We can see that there is a strong positive relationship between a state's population and the number of EV registrations. This is supported by the Pearson correlation coefficient of 0.74 (p < 2.2e-16), suggesting that as population increases, EV registrations tend to rise proportionally. The trendline in the scatterplot confirms this upward trend, indicating that more populous states tend to have significantly higher EV adoption levels—likely due to larger urban areas, better infrastructure, and broader consumer bases.

\newpage

## Analysis 2: Predicting the Number of EV Stations based on EV Sales
### Linear Regression Analysis

